---
title: "HSC"
output: html_document
---

 ---
title: "R Notebook"
o utput:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---
### Loading the packages
```{r}
library(DiffBind)
library(DESeq2)
library(cluster)
library(ggplot2)
library(RColorBrewer)
library(tidyverse)
library(csaw)
library(pheatmap)
library(limma)
library(tidyr)
```

### Read the data
```{r}
library(readr)
ATAC_D = readRDS("ATAC_Seq_groups_all_batches_03_2019_withFileInformation3")
View(ATAC_D)
```


### Make DBA object for diffbind
```{r}
ATAC_D$PeakCaller = "bed"
ATAC_D$Replicate<-"1"
ATAC_D$Condition = ifelse(ATAC_D$age < 60, "Young", "Old")


# ifelse example
# df$newcolumn = ifelse(is.na(df$names), "NA", ifelse(df$favouriteColor =="black", "black", "otherColor"))

ATAC_D$Factor = "ATAC"
names(ATAC_D)[names(ATAC_D) == "peakFile"] <- "Peaks"
names(ATAC_D)[names(ATAC_D) == "bamFile"] <- "bamReads"
#adding a column with gender + sequencing batch that will be later used in design formula for deseq2 analysis
ATAC_D<-unite(ATAC_D,Batch_effect, sex, Batch, remove = FALSE)
View(ATAC_D$Batch_effect)#check if it worked

dba_D= dba(sampleSheet = ATAC_D)
```

### Examine dba object
```{r}
print(dba_D) 
olap.rate = dba.overlap(dba_D, mode = DBA_OLAP_RATE)
plot(olap.rate, main = "overlap rate of HSC peaks") 
plot(dba_D, main = "initial clustering of HSC peaks")
```



```{r}
#generating consensus peaks for ATAC_D
consensusD = dba.count(DBA = dba_D, summits = 250, score = DBA_SCORE_READS, minOverlap = 2)
dds = dba.peakset(consensus, bRetrieve = T, DataType = DBA_DATA_FRAME)
coords = as.data.frame(dds[,1:3], stringsAsFactors = F) # retrieve coordinates
coords$id = paste0(coords[,1], "_", coords[,2], "_", coords[,3])
rownames(dds) = coords$id
colnames(dds) = unlist(sapply(X = colnames(dds), FUN = function(x) gsub(pattern = "X", replacement = "", x = x)))
counts = as.matrix(dds[,4:ncol(dds)])
rownames(ATAC_D) = 1:nrow(ATAC_D)
```

### Run DESeq2
```{r}
dds = DESeqDataSetFromMatrix(counts, colData = ATAC_D, design = ~ Batch_effect + Condition) 
# Batch_effect refers to sex and sequencing batch from the metadata
```

### If the MA plot looks good, proceed directly to differential peak analysis
```{r}
message("\nrunning DESeq2...\n")
dds = DESeq(dds, test = "LRT", reduced=~Batch_effect) # We deviate from the standard workflow here to run likelihood ratio test to account for the different age groups
message("\nDESeq2 completed\n")
res = results(dds, contrast = c("Condition","Young","Old"))   
DESeq2::plotMA(res) # examine distribution of the fold change in dataset

```






### Not necessary if the distribution of log fold change vs mean normalised counts look normal in the MA plot
```{r}

normFacs = exp(normOffsets(counts1, lib.sizes=colSums(counts1), type="loess")) # Loess regression 
normalizationFactors(ddsM1) = normFacs #  run DESeq again using the new normalisation factor

message("\nrunning DESeq2...\n")
ddsM1Norm = DESeq(ddsM1, test = "LRT", reduced=~Batch) # We deviate from the standard workflow here to run likelihood ratio test to account for the different age groups
message("\nDESeq2 completed\n")
plotMA(ddsM1Norm) # examine distribution of the fold change in dataset
```



### We can examine the differential peaks using a heatmap
```{r}
vsd = vst(ddsM1, blind = FALSE) # first transform the data

df <- as.data.frame(colData(ddsM1)[,c("Condition","sex", "age", "SampleID", "Batch")], stringsAsFactors = FALSE)

rownames(df) = paste0(df$SampleID, "_", df$age)
df = df[,c("Batch", "age", "Condition")]
count.matrix <- removeBatchEffect(assay(vsd), batch = colData(ddsM1)$Batch, design = model.matrix( ~ colData(ddsM1)$Condition))

# generate Z-scores from counts
Means.vsd = rowMeans(count.matrix)
SD.vsd = rowSds(count.matrix)
norm.vsd = (count.matrix - Means.vsd) / SD.vsd
colnames(norm.vsd) = rownames(df)
norm.vsd = norm.vsd[,order(sapply(strsplit(colnames(norm.vsd), "_"), "[", 4))]

res = results(dds, contrast = c("Condition", "from_50_to_65", "older_than_65")) #  specify fold change order Old/Young
select_005 = order(res$padj)[1:length(which(res$padj < 0.05))] # choose p adjusted value as the FDR cut off

x = norm.vsd[select_005,]
clusters = pam(x, 2) # use pam clustering here to obtain consistent clusters, you can choose the cluster number 
x = x[names(sort(clusters$clustering)),]

pheatmap(x, cluster_rows = FALSE,
         cluster_cols = FALSE,
         show_rownames=FALSE,
         show_colnames = T,
         annotation_col=df)
```

### We can examine the peaks using a PCA plot
```{r}
cct = assay(vsd, blind  = T)
pvars <- rowVars(cct)
select <- order(pvars, decreasing = TRUE)[seq_len(min(500, length(pvars)))] #  select top 500 peaks
PCA <- prcomp(t(cct)[, select], scale = F)
percentVar <- round(100 * PCA$sdev ^ 2/ sum(PCA$sdev ^ 2), 1)
plotPCA.df = data.frame(PC1 = PCA$x[,1], PC2 = PCA$x[,2], PC3 = PCA$x[,3], PC4 = PCA$x[,4], sample = paste0(colData(ddsM1)$SampleID,"_", colData(ddsM1)$age), condition = as.character(colData(ddsM1)$age), batch = colData(dds)$Batch)

qplot(PC1, PC2, data = plotPCA.df, color =  Condition, shape = batch, main = "PC1vsPC2 top 500 peaks", size = I(6))+ labs(x = paste0("PC1, VarExp:", round(percentVar[1],4)),  y = paste0("PC2, VarExp:", round(percentVar[2],4))) + scale_color_manual(values = colorRampPalette(brewer.pal(n = 5, name = "RdBu"))(16))+ theme_classic()



```

#